State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-05-09 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
edits to discussion. added more to first paragraph

% \address{$^1$ Department of Electrical and Electronic Engineering, University of Melbourne, Melbourne, Vic, 3010 Australia} \address{$^2$ Department of Automatic Control and Systems Engineering, University of Sheffield, Mappin Street, Sheffield, S1 3JD, UK} \address{$^3$ Department of Applied Physics and Applied Mathematics, Columbia University, US} \address{$^4$ Ken's} \ead{dfreestone@bionicear.org} 
\begin{abstract}
	Will do the abstract towards the end. 
\end{abstract}

@@

\subsection{Derivation of Finite Dimensional State-Space Model} 
In order to implement standard estimation techniques we use a decomposition of the field using a set of Gaussian basis functions defined by
\begin{equation}\label{eq:FieldBasisFunction}
	\phi\left(\mathbf{r}-\mathbf{r}'\right) =
\exp{\left(-\frac{(\mathbf{r}-\mathbf{r}')^\top(\mathbf{r}-\mathbf{r}')}{\sigma_{\phi}^2}\right)}. 
\end{equation}
 Decomposition allows a continuous field to be represented by a finite dimensional state vector. This facilitates application of standard nonlinear state estimation methods such as the unscented Kalman filter. The field decomposition is described by 
\begin{equation}
	\label{DefFieldDecomp} v_t\left(\mathbf{r}\right) \approx \boldsymbol{\phi}^{\top}\left(\mathbf{r}\right) \mathbf{x}_t, 
\end{equation}
where $\mathbf{\boldsymbol{\phi}}(\mathbf{r})$ is a vector of Gaussian basis functions that are scaled by the state vector, $\mathbf{x}_t$. The width and positioning of the basis function can be determined by spectral analysis explain in detail in Section~\ref{SpectralAnalysisSection}. The connectivity kernel can also be considered a decomposition into a set of basis functions in 
\begin{equation}\label{DefKernelDecomp}
	 w\left(\mathbf{r}-\mathbf{r}'\right) =\boldsymbol{\psi}^\top\left(\mathbf{r}-\mathbf{r}'\right) \boldsymbol{\theta}.
\end{equation}
We will assume we know the parametric form of the connectivity basis functions, where the parameter $\boldsymbol{\theta}$ is unknown. Each connectivity basis function can, individually, be considered a layer in the Wilson and Cowan model, representing short range excitation, surround inhibition and mid-range excitation. Making substitutions of equations~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} into~\ref{DiscreteTimeModel} we get 
\begin{eqnarray}
	\label{reduced continuous model}\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_{t+1}&=& T_s\int_\Omega{f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t )\boldsymbol{\psi}^{\top}(\mathbf{r}-\mathbf{r}')d\mathbf{r}'}\boldsymbol{\theta}\nonumber \\ 
	&-& \xi\boldsymbol{\phi}^{\top}(\mathbf{r})x_t + e_t(\mathbf{r}). 
\end{eqnarray}
By defining the term 
\begin{equation}
	\label{DefGamma} \boldsymbol{\Gamma} = \int_\Omega {\boldsymbol{\phi} \left(\mathbf{r}\right)\boldsymbol{\phi} ^{\top}\left(\mathbf{r}\right)d\mathbf{r}} 
\end{equation}
and manipulating the model (see Appendix~\ref{Simplifying Decomposition}) we isolate the state vector to form a state-space model 
\begin{eqnarray}
	\label{Homogeneous SS Model} \mathbf{x}_{t+1} &=& q(\mathbf{x}_t)\boldsymbol{\theta} - \xi\mathbf{x}_t + \mathbf{e}_t \\
	&=& Q\left(\mathbf{x}_t\right) + \mathbf{e}_t 
\end{eqnarray}
where
\begin{eqnarray}
Q\left( \mathbf{x}_t \right) &=& q(\mathbf{x}_t)\boldsymbol{\theta} - \xi\mathbf{x}_t \label{eq:QmatrixForSigmapoints}\\
	\mathbf{e}_t &=& \label{eq:Reducednoiseterm} \boldsymbol{\Gamma}^{-1}\int_\Omega{\boldsymbol{\phi}(\mathbf{r})e_t(\mathbf{r})d\mathbf{r}}, 
\end{eqnarray}
where $\mathbf{e}_t$ is a zero-mean normal white noise term with the covariance (see~\ref{ColoredNoise}) 
\begin{equation}
	\boldsymbol{\Sigma}_e=\boldsymbol{\Gamma}^{-1}\int_{\Omega}\int_{\Omega}\boldsymbol{\phi}\left(\mathbf r\right) \gamma\left(\mathbf{r}- \mathbf{r}' \right)\boldsymbol{\phi}\left(\mathbf{r}'\right)^{\top}d\mathbf{r}' d\mathbf{r}\boldsymbol{\Gamma}^{- \top} 
\end{equation}
The observation equation of the reduced model is 
\begin{equation}
	\label{ObservationEquation} \mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\varepsilon}_t 
\end{equation}
where the observation matrix is 
\begin{equation}
	\mathbf{C} = \left[
	\begin{array}{{ccc}} 
		c_{1,1} & \dots & c_{1,L} \\
		\vdots & \ddots & \vdots \\
		c_{n_y,L} & \dots & c_{n_y,L} 
	\end{array}
	\right] 
\end{equation}
where $L$ is the number of basis functions and 
\begin{equation}
	c_{i,j} = \int_{\Omega}m(\mathbf{r}_i - \mathbf{r}')\boldsymbol{\phi}_j(\mathbf{r}')d\mathbf{r}'. 
\end{equation}

@@

The Unscented Rauch Tung Striebel Smoother (URTSS)~\cite{Sarkka2010} was used for the state estimation. The URTSS uses the unscented Kalman Filter (UKF)~\cite{Julier1997, Merwe2003} in a forward iteration to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward iteration to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and the second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. The sigma points, $\mathcal X_i$, are calculated using the unscented transform as follows

@@

\omg{FIRST DESCRIBE AN OVER ALL SUMMARY OF WHAT WE HAVE DONE AND WHAT IS NOVEL}
This paper presents a novel framework for model reduction to facilitate practical state and parameter estimation of neural fields. The level of detail of the estimated field is related to the basis functions decomposition, thereby providing a flexible approach where a trade-off can be made between physiological detail and computational complexity. The choice of Gaussian basis functions can be justified by the existence of the so called bump solutions for this class of model~\cite{Coombes2005}.

@@

\omg{NEED TO ADD SOMETHING ABOUT THE TIME CONSTANT}

>>>>

\address{$^1$ Department of Electrical and Electronic Engineering, University of Melbourne, Melbourne, Vic, 3010 Australia} \address{$^2$ Department of Automatic Control and Systems Engineering, University of Sheffield, Mappin Street, Sheffield, S1 3JD, UK} \address{$^3$ Department of Applied Physics and Applied Mathematics, Columbia University, US} \address{$^4$ Ken's} \ead{dfreestone@bionicear.org} 
\begin{abstract}
	Will do the abstract towards the end. 
\end{abstract}

@@

\subsection{Derivation of Finite Dimensional State-Space Model} 
In order to implement standard estimation techniques we use a decomposition of the field using a set of Gaussian basis functions defined by
\begin{equation}\label{eq:FieldBasisFunction}
	\phi\left(\mathbf{r}-\mathbf{r}'\right) =
\exp{\left(-\frac{(\mathbf{r}-\mathbf{r}')^\top(\mathbf{r}-\mathbf{r}')}{\sigma_{\phi}^2}\right)}. 
 Decomposition allows a continuous field to be represented by a finite dimensional state vector. This facilitates application of standard nonlinear state estimation methods such as the unscented Kalman filter. The field decomposition is described by 
	\label{DefFieldDecomp} v_t\left(\mathbf{r}\right) \approx \boldsymbol{\phi}^{\top}\left(\mathbf{r}\right) \mathbf{x}_t, 
where $\mathbf{\boldsymbol{\phi}}(\mathbf{r})$ is a vector of Gaussian basis functions that are scaled by the state vector, $\mathbf{x}_t$. The choice of Gaussian basis functions can be justified by the existence of the so called bump solutions for this class of model~\cite{Coombes2005}. The width and positioning of the basis function can be determined by spectral analysis explain in detail in Section~\ref{SpectralAnalysisSection}. The connectivity kernel can also be considered a decomposition into a set of basis functions in 
\begin{equation}\label{DefKernelDecomp}
	 w\left(\mathbf{r}-\mathbf{r}'\right) =\boldsymbol{\psi}^\top\left(\mathbf{r}-\mathbf{r}'\right) \boldsymbol{\theta}.
We will assume we know the parametric form of the connectivity basis functions, where the parameter $\boldsymbol{\theta}$ is unknown. Each connectivity basis function can, individually, be considered a layer in the Wilson and Cowan model, representing short range excitation, surround inhibition and mid-range excitation. Making substitutions of equations~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} into~\ref{DiscreteTimeModel} we get 
	\label{reduced continuous model}\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_{t+1}&=& T_s\int_\Omega{f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t )\boldsymbol{\psi}^{\top}(\mathbf{r}-\mathbf{r}')d\mathbf{r}'}\boldsymbol{\theta}\nonumber \\ 
	&-& \xi\boldsymbol{\phi}^{\top}(\mathbf{r})x_t + e_t(\mathbf{r}). 
By defining the term 
	\label{DefGamma} \boldsymbol{\Gamma} = \int_\Omega {\boldsymbol{\phi} \left(\mathbf{r}\right)\boldsymbol{\phi} ^{\top}\left(\mathbf{r}\right)d\mathbf{r}} 
and manipulating the model (see Appendix~\ref{Simplifying Decomposition}) we isolate the state vector to form a state-space model 
	\label{Homogeneous SS Model} \mathbf{x}_{t+1} &=& q(\mathbf{x}_t)\boldsymbol{\theta} - \xi\mathbf{x}_t + \mathbf{e}_t \\
	&=& Q\left(\mathbf{x}_t\right) + \mathbf{e}_t 
Q\left( \mathbf{x}_t \right) &=& q(\mathbf{x}_t)\boldsymbol{\theta} - \xi\mathbf{x}_t \label{eq:QmatrixForSigmapoints}\\
	\mathbf{e}_t &=& \label{eq:Reducednoiseterm} \boldsymbol{\Gamma}^{-1}\int_\Omega{\boldsymbol{\phi}(\mathbf{r})e_t(\mathbf{r})d\mathbf{r}}, 
where $\mathbf{e}_t$ is a zero-mean normal white noise term with the covariance (see~\ref{ColoredNoise}) 
	\boldsymbol{\Sigma}_e=\boldsymbol{\Gamma}^{-1}\int_{\Omega}\int_{\Omega}\boldsymbol{\phi}\left(\mathbf r\right) \gamma\left(\mathbf{r}- \mathbf{r}' \right)\boldsymbol{\phi}\left(\mathbf{r}'\right)^{\top}d\mathbf{r}' d\mathbf{r}\boldsymbol{\Gamma}^{- \top} 
The observation equation of the reduced model is 
	\label{ObservationEquation} \mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\varepsilon}_t 
where the observation matrix is 
	\mathbf{C} = \left[
	\begin{array}{{ccc}} 
		c_{1,1} & \dots & c_{1,L} \\
		\vdots & \ddots & \vdots \\
		c_{n_y,L} & \dots & c_{n_y,L} 
where $L$ is the number of basis functions and 
	c_{i,j} = \int_{\Omega}m(\mathbf{r}_i - \mathbf{r}')\boldsymbol{\phi}_j(\mathbf{r}')d\mathbf{r}'. 

@@

The unscented Rauch Tung Striebel Smoother (URTSS)~\cite{Sarkka2010} was used for the state estimation. The URTSS uses the unscented Kalman Filter (UKF)~\cite{Julier1997, Merwe2003} in a forward iteration to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward iteration to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and the second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. The sigma points, $\mathcal X_i$, are calculated using the unscented transform as follows

@@

\omg{FIRST DESCRIBE AN OVER ALL SUMMARY OF WHAT WE HAVE DONE AND WHAT IS NOVEL}
This paper presents a novel framework for model reduction to facilitate practical state and parameter estimation of neural fields. A basis function decomposition reduces the infinite dimensional neural field (continuous space), to a finite dimensional state-space model. The level of detail of the estimated field is related to the basis functions decomposition, thereby providing a flexible approach where a trade-off can be made between physiological detail and computational complexity. By forming a state-space model, the unscented Rauch Tung Striebel Smoother was implemented for state estimation. Using the sequence of state estimates, a least squares algorithm was implemented for parameter estimation.

@@

\omg{NEED TO ADD SOMETHING ABOUT THE TIME CONSTANT }
